---
title: "Forecasting with Multiple Time Series"
author: "Nickalus Redell"
date: "`r lubridate::today()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Forecasting with Multiple Time Series}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 4)
```

## Purpose

The purpose of this vignette is to provide a quick overview of forecasting with multiple time-series in 
`forecastML`. The benefits to modeling multiple time-series in one go with a single model or ensemble of 
models include (a) modeling simplicity, (b) potentially more robust results from pooling data across time-series, 
and (c) solving the cold-start problem when few data points are available for a given time-series.

We won't concern ourselves with error metrics or hyperparameter evaluation which can be assessed with the 
`return_error()` and `return_hyper()` functions. This functionality is covered 
in the package overview vignette.

## Setup

To forecast with multiple/grouped/hierarchical time-series in `forecastML`, your data needs the following 
characteristics:

* The **same outcome** is being forecasted across time-series.

* Data are in a **long format** with a single outcome column--i.e., time-series are stacked on top of each other 
in a data.frame.

* There are 1 or more **grouping columns**.

* There **may be 1 or more static features** that are constant through time but differ between time-series--e.g., 
a fixed location, store square footage, species of animal etc.

* The **time-series are regularly spaced** and have no missing rows or gaps in time. Irregular or sparse time-series 
with many `NA`s *can* be modeled in this framework, but missing rows will result in incorrect feature 
lags when using `create_lagged_df()` which is the first step in the `forecastML` workflow. To fix any 
gaps in data collection, use the `fill_gaps()` function; handling the resulting missing values in the 
target being forecasted and the dynamic features can be done (a) prior to `create_lagged_df()` or (b) 
in the user-defined model training function.

## Example

To illustrate forecasting with multiple time-series, we'll use the `data_buoy` dataset that comes 
with the `forecastML` package. This dataset consists of daily sensor measurements of several environmental 
conditions collected by 14 buoys in Lake Michigan from 2012 through 2018. The data were obtained 
from NOAA's National Buoy Data Center available at https://www.ndbc.noaa.gov/ using the `rnoaa` package.

* **Outcome:** Average daily wind speed in Lake Michigan.

* **Forecast horizon:** Daily, 1 to 30 days into the future which is essentially January 2019 for this dataset.

* **Time-series:** 14 outcome time-series collected from buoys throughout Lake Michigan.

* **Model:** A single gradient boosted tree model with `xgboost`.

### Load packages and data

`data_buoy_gaps` consists of:

* `date`: A date column which will be removed for modeling.

* `wind_spd`: The outcome.

* `lat` and `lon`: Latitude and longitude which are features that are static or unchanging through time.

* `day` and `year`: Dynamic features encoding time.

* `air_temperature` and `sea_surface_temperature`: Data collected from the buoys through time.


```{r, message = FALSE, warning = FALSE}
library(dplyr)
library(DT)
library(ggplot2)
library(forecastML)
library(xgboost)

data("data_buoy_gaps", package = "forecastML")

DT::datatable(head(data_buoy_gaps), options = list(scrollX = TRUE))
```

### forecastML::fill_gaps

* The wind speed data has some gaps in it: Some buoys collected data throughout the year, others only 
during the summer months. These gaps in data collection would result in incorrect feature lags in 
`create_lagged_df()` as the previous row in the dataset for a given buoy--a lag of 1--may be several 
months in the past.

* To fix this problem, we'll run `fill_gaps()` to fill in the rows for the missing dates. The added rows will appear 
between `min(date)` for each buoy and and `max(date)` across all buoys. For example, buoy 45186 that only started data collection in 2018 won't 
have additional rows with `NA`s for 2012 through 2017; only gaps since the start of data collection in 2018 to the most recent date will be filled.

* After running `fill_gaps()`, **the following columns have been filled in and have no `NA`s**: `date`, `buoy_id`, `lat`, and `lon`.

* After running `fill_gaps()`, **the following columns now have additional `NA`s**: our `wind_spd` target and the dynamic features.

* Notice that the input dataset and the returned dataset have the same columns in the same order with the same data types.

```{r}
data <- forecastML::fill_gaps(data_buoy_gaps, date_col = 1, frequency = '1 day', 
                              groups = 'buoy_id', static_features = c('lat', 'lon'))

print(list(paste0("The original dataset with gaps in data collection is ", nrow(data_buoy_gaps), " rows."), 
      paste0("The modified dataset with no gaps in data collection from fill_gaps() is ", nrow(data), " rows.")))
```


```{r}
DT::datatable(head(data), options = list(scrollX = TRUE))
```

### Dynamic features

* Now would be a good time to fill in the newly created missing values in our dynamic features: day and year. These 
features are deterministic and won't be lagged in the modeling dataset. We could also impute missing values for 
air_temperature and sea_surface_temperature, but we'll let our `xgboost` model handle these `NA`s.

```{r}
data$day <- lubridate::mday(data$date)
data$year <- lubridate::year(data$date)
```


### Plot wind speed outcome

* Notice that buoy 45186 has only recently come online and would be difficult to forecast on its own.

```{r, message = FALSE, warning = FALSE}
p <- ggplot(data, aes(x = date, y = wind_spd, color = ordered(buoy_id), group = year))
p <- p + geom_line()
p <- p + facet_wrap(~ ordered(buoy_id), scales = "fixed")
p <- p + theme_bw() + theme(
  legend.position = "none"
) + xlab(NULL)
p
```

### Modeling setup

* We'll simply and incorrectly set our grouping column, `buoy_id`, to numeric to work smoothly with `xgboost`. 
Better alternatives include [feature embedding](https://arxiv.org/abs/1604.06737), 
[target encoding](http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-munging/target-encoding.html) 
(available in the `R` package `catboost`), or [mixed effects Random Forests](https://arxiv.org/pdf/1901.11279.pdf).

* To be clear, `buoy_id` is both (a) used to identify a specific time-series for creating lagged features 
and (b) used as a feature in the model.

```{r}
data$buoy_id <- as.numeric(factor(data$buoy_id))
```

<br>

* Inputs for dataset creation with `create_lagged_df()`.

```{r}
outcome_col <- 1  # The column position of our 'wind_spd' outcome.

horizons <- c(1, 7, 30)  # Forecast 1, 1:7, and 1:30 days into the future.

lookback <- c(1:30, 360:370)  # Features from 1 to 30 days in the past and annually.

dates <- data$date  # Grouped time series forecasting requires dates.
data$date <- NULL  # Dates, however, don't need to be in the input data.

frequency <- "1 day"  # A string that works in base::seq(..., by = "frequency").

dynamic_features <- c("day", "year")  # Features that change through time but which we will not lag.

groups <- "buoy_id"  # 1 forecast for each group or buoy.

static_features <- c("lat", "lon")  # Features that do not change through time.
```

<br>

### Model training with nested CV

#### Training dataset

* We have 3 datasets for training models that forecast 1, 1 to 7, and 1 to 30 days into 
the future. We'll view the 1-day-ahead training data below.

```{r}
type <- "train"  # Create a model-training dataset.

data_train <- forecastML::create_lagged_df(data, type = type, outcome_col = outcome_col,
                                           horizons = horizons, lookback = lookback,
                                           dates = dates, frequency = frequency,
                                           dynamic_features = dynamic_features,
                                           groups = groups, static_features = static_features, 
                                           use_future = FALSE)

print(paste0("The class of `data_train` is ", class(data_train)))

DT::datatable(head(data_train$horizon_1), options = list(scrollX = TRUE))
```

<br>

* The plot below shows the feature map across forecast horizons. Here, we set all 
non-dynamic/static features to have the same lags (refer to the custom lags vignette to see how this could be modified). 
Notice that features that don't support direct forecasting to the given horizon are silently dropped.

```{r, message = FALSE, warning = FALSE}
p <- plot(data_train)  # plot.lagged_df() returns a ggplot object.
p <- p + geom_tile(NULL)  # Remove the gray border for a cleaner plot.
p
```

#### CV setup

* We'll model with **3 validation datasets**. 
Given that our measurements are taken daily, we'll set the `skip = 730` argument to skip 2 years 
between validation datasets.

```{r, message = FALSE, warning = FALSE}
windows <- forecastML::create_windows(data_train, window_length = 365, skip = 730,
                                      include_partial_window = FALSE)

p <- plot(windows, data_train) + theme(legend.position = "none")
p
```

* Now we'll use the `group_filter = "buoy_id == 1"` argument to get a closer look at 
1 of our 14 time-series. The user-supplied filter is passed to `dplyr::filter()` internally.

```{r, message = FALSE, warning = FALSE}
p <- plot(windows, data_train, group_filter = "buoy_id == 1") + 
  theme(legend.position = "none")
p
```

#### User-defined modeling function

* A user-defined wrapper function for model training that takes the following arguments: 
    + **1:** A horizon-specific data.frame made with `create_lagged_df(..., type = "train")` 
    (e.g., my_lagged_df$horizon_h),
    + **2:** optionally, any number of additional named arguments which can be passed as '...' in
    `train_model()`
    + and **returns** a model object that will be passed into the user-defined `predict()` function.

Any data transformations, hyperparameter tuning, or inner loop cross-validation procedures should take 
place within this function, with the limitation that it ultimately needs to `return()` a model suitable for 
the user-defined `predict()` function; a list can be returned to capture meta-data such as hyperparameter results.

* Notice that the `xgboost`-specific input datasets are created within this wrapper function.

```{r}
# The value of outcome_col can also be set in train_model() with train_model(outcome_col = 1).
model_function <- function(data, outcome_col = 1) {
  
  # xgboost cannot handle missing outcomes data so we'll remove it.
  data <- data[!is.na(data[, outcome_col]), ]

  # 1 fixed validation dataset for early stopping. Ideally, we'd have many. 
  # We'll use an 80/20 split.
  indices <- 1:nrow(data)
  
  set.seed(224)
  train_indices <- sample(1:nrow(data), ceiling(nrow(data) * .8), replace = FALSE)
  test_indices <- indices[!(indices %in% train_indices)]

  data_train <- xgboost::xgb.DMatrix(data = as.matrix(data[train_indices, 
                                                           -(outcome_col), drop = FALSE]),
                                     label = as.matrix(data[train_indices, 
                                                            outcome_col, drop = FALSE]))

  data_test <- xgboost::xgb.DMatrix(data = as.matrix(data[test_indices, 
                                                          -(outcome_col), drop = FALSE]),
                                    label = as.matrix(data[test_indices, 
                                                           outcome_col, drop = FALSE]))

  params <- list("objective" = "reg:linear")

  watchlist <- list(train = data_train, test = data_test)
  set.seed(224)
  model <- xgboost::xgb.train(data = data_train, params = params, 
                              max.depth = 8, nthread = 2, nrounds = 30,
                              metrics = "rmse", verbose = 0, 
                              early_stopping_rounds = 5, 
                              watchlist = watchlist)

  return(model)
}
```

<br>

#### Model training

* This should take ~1 minute to train our '3 forecast horizons' * '3 validation datasets' = *9 models*.

* The user-defined modeling wrapper function could be much more elaborate, in which case many more models 
could potentially be trained here.

* These models could be trained in parallel on any OS with the very flexible `future` package by uncommenting the code below and 
setting `use_future = TRUE`. To avoid nested parallelization, models are either trained in parallel across forecast horizons or 
validation windows, whichever is longer (when equal, the default is parallel across forecast horizons).

```{r}
#future::plan(future::multiprocess)  # Multi-core or multi-session parallel training.

model_results_cv <- forecastML::train_model(lagged_df = data_train,
                                            windows = windows,
                                            model_name = "xgboost",
                                            model_function = model_function, 
                                            use_future = FALSE)
```


```{r}
print(paste0("The class of `model_results_cv` is ", class(model_results_cv)))
```

<br>

* We can access the `xgboost` model for any horizon or validation window. Here, 
we show a `summary()` of the model for the first validation window which is 2012.

```{r}
summary(model_results_cv$horizon_1$window_1$model)
```

### Forecasting with nested models

* First, we'll visually evaluate our model's performance across our validation datasets.

* Then we'll forecast with each of our 9 models to get a sense of the stability of the forecasts 
produced from models trained on different subsets of our historical data.

#### User-defined prediction function

The following user-defined prediction function is needed for each model:

* A wrapper function that takes the following **2 positional arguments**:
    * **1:** The model returned from the user-defined modeling function.
    * **2:** A `data.frame` of the model features from `forecastML::create_lagged_df(..., type = "train")`.
* and **returns** a `data.frame` of predictions with 1 or 3 columns. A 1-column data.frame will produce point forecasts, 
and a 3-column data.frame can be used to return point, lower, and upper forecasts (column names and order do not matter).

```{r}
prediction_function <- function(model, data_features) {
  x <- xgboost::xgb.DMatrix(data = as.matrix(data_features))
  data_pred <- data.frame("y_pred" = predict(model, x))
  return(data_pred)
}

prediction_function <- list(prediction_function)
```

<br>

#### Historical model fit

* Herer, we're predicting on our validation datasets.

```{r}
data_pred_cv <- predict(model_results_cv, prediction_function = prediction_function, data = data_train)

print(paste0("The class of `data_pred_cv` is ", class(data_pred_cv)))
```

* The predictions are solid lines; the actuals are dashed lines.

* We'll filter this plot for closer inspection below.

```{r, message = FALSE, warning = FALSE}
plot(data_pred_cv) + theme(legend.position = "none")
```

<br>

* It's somewhat difficult to see how we've done here, so we'll use the `group_filter` argument again 
to focus on specific results. Note that the `plot()` function returns a `ggplot` object so that we 
can easily modify our plot.

* Notice that during the first half of 2015 there were no wind speed measurements recorded; however, 
we're getting some interesting variability in predictions here because the day of the month and year dynamic features 
have captured some seasonality.

```{r, message = FALSE, warning = FALSE}
p <- plot(data_pred_cv, group_filter = "buoy_id %in% c(1, 2, 3)", horizons = 7) 
p <- p + theme(legend.position = "none")
p <- p + scale_x_date(limits = c(as.Date("2015-01-01"), as.Date("2015-12-31")))
p <- p + facet_grid(horizon ~ buoy_id) + ggtitle(paste0(p$labels$title, " and buoy ID"))
p <- p + theme(axis.text.x = element_text(angle = 90))
p
```

#### Forecasting

* We have 3 datasets that support forecasting 1, 1 to 7, and 1 to 30 days into 
the future. We'll view the 1-day-ahead forecasting data below.

```{r}
type <- "forecast"  # Create a forecasting dataset for our predict() function.

data_forecast <- forecastML::create_lagged_df(data, type = type, outcome_col = outcome_col,
                                              horizons = horizons, lookback = lookback,
                                              dates = dates, frequency = frequency,
                                              dynamic_features = dynamic_features,
                                              groups = groups, static_features = static_features, 
                                              use_future = FALSE)

print(paste0("The class of `data_forecast` is ", class(data_forecast)))

DT::datatable(head(data_forecast$horizon_1), options = list(scrollX = TRUE))
```

#### Dynamic features and forecasting

* Our dynamic features 'day' and 'year' were not lagged in our modeling dataset. This was the right choice 
from a modeling perspective; however, in order to forecast 'h' steps ahead, we need to know their future 
values for each forecast horizon. At present, there's no function in `forecastML` to autofill the future 
values of dynamic, non-lagged features so we'll simply do it manually below.

```{r}
for (i in seq_along(data_forecast)) {
  data_forecast[[i]]$day <- lubridate::mday(data_forecast[[i]]$index)
  data_forecast[[i]]$year <- lubridate::year(data_forecast[[i]]$index)
}
```

#### Forecast

* Now we'll forecast 1, 1:7, and 1:30 days into the future with `predict(..., data = data_forecast)`.

* The first time step into the future is `max(dates) + 1 * frequency`. Here, this is 
12-31-2018 + 1 * '1 day' or 1-1-2019.

```{r}
data_forecasts <- predict(model_results_cv, prediction_function = prediction_function, 
                          data = data_forecast)

print(paste0("The class of `data_forecasts` is ", class(data_forecasts)))
```

* Plots for each model--just `xgboost` here--and time horizon. The separate lines are for the 
combination of `buoy_id` * validation window.

```{r, message = FALSE, warning = FALSE}
plot(data_forecasts) + theme(legend.position = "none")
```

* We'll filter results to look at the forecasts across validation windows for `buoy_id == 1`. 
In this case, a validation window shows how the model would forecast if it were not trained 
on data in a given time frame.

```{r, message = FALSE, warning = FALSE}
plot(data_forecasts, group_filter = "buoy_id == 1")
```

### Model training all data

* The modeling steps are more or less the same as in the nested cross-validation modeling described 
above so we'll skip the explanations from here on out.

* Notice that by this point in the modeling process, the optimal hyperparameters that gave the best performance on 
the outer-loop validation datasets have already been identified (see the package overview vignette). Incorporating the 
optimal hyperparameters in a final model would occur in a new user-defined modeling wrapper function.

* Train across all data by setting `window_length = 0`.

```{r, message = FALSE, warning = FALSE}
windows <- forecastML::create_windows(data_train, window_length = 0)

p <- plot(windows, data_train) + theme(legend.position = "none")
p
```


```{r}
# Un-comment the code below and set 'use_future' to TRUE.
#future::plan(future::multiprocess)

model_results_no_cv <- forecastML::train_model(lagged_df = data_train, 
                                               windows = windows,
                                               model_name = "xgboost",
                                               model_function = model_function,
                                               use_future = FALSE)
```


```{r}
print(paste0("The class of `model_results_no_cv` is ", class(model_results_no_cv)))
```


```{r}
data_forecasts <- predict(model_results_no_cv, prediction_function = prediction_function, 
                          data = data_forecast)

print(paste0("The class of `data_forecasts` is ", class(data_forecasts)))
```


```{r}
DT::datatable(head(data_forecasts), options = list(scrollX = TRUE))
```


```{r, message = FALSE, warning = FALSE}
plot(data_forecasts)
```


```{r, message = FALSE, warning = FALSE}
plot(data_forecasts, group_filter = "buoy_id == 1") + theme(legend.position = "none")
```
